---
title: "DeNovo_Motifs_DiffMeth"
author: "Michael Ruesch"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Designate the library path within the the rnbeads_2.24.0 conda environment
```{r}
.libPaths("/users/PAS1800/ruesch6/mambaforge/envs/rnbeads_2.24.0/lib/R/library")
```

Load Packages
```{r include=FALSE}
library(RnBeads)
library(RnBeads.hg38)
library(ggrepel)
library(dplyr)
library(tidyr)
library(tibble)
```

Load in RnBeads Set and Confirm Phenotype of Samples
Add in CellType value in phenotype to help with analysis
```{r}
NK_RnB <- load.rnb.set("~/DNAmethylation/RnBeads_Output/NKDI_and_IV_RnBeadsSet/reports/rnbSet_preprocessed")
phenotype <- pheno(NK_RnB)
phenotype$CellType <- c(rep("15D_D14", 2), 
                        rep("15C_D14", 2), 
                        rep(c("15D_CD16Neg", "15D_CD16Pos"),2),
                        rep(c("15C_CD16Neg", "15C_CD16Pos"), 2),
                        rep("15C_D14", 2),
                        rep(c("15D_CD16Neg", "15D_CD16Pos"),2),
                        "15C_CD16Neg", "15C_CD16Pos", "15C_CD16Neg",
                        rep("S1",4), 
                        rep("S2A", 4),
                        rep("S2B", 4),
                        rep("S3",4), 
                        rep("S4A", 4),
                        rep("S4B", 4),
                        rep("S5", 3)
)
slot(NK_RnB, "pheno") <- phenotype
```

Subset on just the CD15C and 15D CD16- NK cells from day 28
Also doing CD16+ NK Cells as well
As well as the Day14 NK cells from each condition
```{r}
# Find indices to remove
remove_idx <- which(!pheno(NK_RnB)$CellType %in% c("15D_CD16Neg", "15C_CD16Neg"))
# Remove them
NK_RnB_subset <- remove.samples(NK_RnB, remove_idx)
# Check
table(pheno(NK_RnB_subset)$CellType)
#RunDiffMeth on just the CD15C and 15D CD16- NK Cells
#CD16Neg_diffMeth <- rnb.execute.computeDiffMeth(NK_RnB_subset, 
#                                        pheno.cols = "CellType", 
#                                        pheno.cols.all.pairwise = T)

# Find indices to remove
remove_idx <- which(!pheno(NK_RnB)$CellType %in% c("15D_CD16Pos", "15C_CD16Pos"))
# Remove them
NK_RnB_subset2 <- remove.samples(NK_RnB, remove_idx)
# Check
table(pheno(NK_RnB_subset2)$CellType)
#RunDiffMeth on just the CD15C and 15D CD16- NK Cells
#CD16Pos_diffMeth <- rnb.execute.computeDiffMeth(NK_RnB_subset2, 
#                                        pheno.cols = "CellType", 
#                                        pheno.cols.all.pairwise = T)

# Find indices to remove
remove_idx <- which(!pheno(NK_RnB)$CellType %in% c("15D_D14", "15C_D14"))
# Remove them
NK_RnB_subset3 <- remove.samples(NK_RnB, remove_idx)
# Check
table(pheno(NK_RnB_subset3)$CellType)
#RunDiffMeth on just the CD15C and 15D CD16- NK Cells
D14_diffMeth <- rnb.execute.computeDiffMeth(NK_RnB_subset3, 
                                        pheno.cols = "CellType", 
                                        pheno.cols.all.pairwise = T)

#Save RnBeads DiffMeth Files
#save.rnb.diffmeth(CD16Neg_diffMeth, "~/DNAmethylation/NKDI_NK_IV_Analysis/De_Novo_Motifs_Homer/CD16Neg_diffMeth")
CD16Neg_diffMeth <- load.rnb.diffmeth("~/DNAmethylation/NKDI_NK_IV_Analysis/De_Novo_Motifs_Homer/CD16Neg_diffMeth")
#save.rnb.diffmeth(CD16Pos_diffMeth, "~/DNAmethylation/NKDI_NK_IV_Analysis/De_Novo_Motifs_Homer/CD16Pos_diffMeth")
CD16Pos_diffMeth <- load.rnb.diffmeth("~/DNAmethylation/NKDI_NK_IV_Analysis/De_Novo_Motifs_Homer/CD16Pos_diffMeth")
#save.rnb.diffmeth(D14_diffMeth, "~/DNAmethylation/NKDI_NK_IV_Analysis/De_Novo_Motifs_Homer/D14_diffMeth")
D14_diffMeth <- load.rnb.diffmeth("~/DNAmethylation/NKDI_NK_IV_Analysis/De_Novo_Motifs_Homer/D14_diffMeth")
```

Find the hypo and hyper methylated probes for each comparison
```{r}
#Start with CD16- NK cells
s.groups <- rnb.sample.groups(NK_RnB_subset,"CellType")[[1]] #Pull comparison group
dmt.sites <- computeDiffTab.extended.site(meth(NK_RnB_subset),s.groups[[1]],s.groups[[2]], paired = T) #Compute differential methylation in this condition for each probe
meth_CD16Neg <- meth(NK_RnB_subset, type = "sites", row.names = TRUE) #Get the beta values for the CD16- NK 
rownames(dmt.sites) <- rownames(meth_CD16Neg) #Apply the probe rownames to the 
pref_probes <- read.csv("~/DNAmethylation/NKDI_NK_IV_Analysis/805KAnnotation.csv") #Call in preferred probes for EPIC v1 based on Oakes lab filtering
filtered_dmt <- dmt.sites[rownames(dmt.sites) %in% pref_probes$IlmnID, ] #filter on only the preferred probes
filtered_dmt <- filtered_dmt[complete.cases(filtered_dmt), ] #remove any NA values
hyper_probes <- rownames(filtered_dmt[filtered_dmt$mean.diff > 0.2, ]) #Comparison is 15C vs 15D so 15C Beta minus 15D Beta
hypo_probes <- rownames(filtered_dmt[filtered_dmt$mean.diff < -0.2, ])
probe_anno <- annotation(NK_RnB_subset, type = "sites", add.names = T) #Call the annotation information for each probe from RnBeads
CD16Neg_hyper <- probe_anno[rownames(probe_anno) %in% hyper_probes, ] #Filter on just the hyper methylated probes
CD16Neg_hypo <- probe_anno[rownames(probe_anno) %in% hypo_probes, ] #Filter on just the hypo methylated probes

#Next with CD16+ NK cells
s.groups <- rnb.sample.groups(NK_RnB_subset2,"CellType")[[1]] #Pull comparison group
dmt.sites <- computeDiffTab.extended.site(meth(NK_RnB_subset2),s.groups[[1]],s.groups[[2]], paired = F) #Compute differential methylation in this condition for each probe
meth_CD16Pos <- meth(NK_RnB_subset2, type = "sites", row.names = TRUE) #Get the beta values for the CD16- NK 
rownames(dmt.sites) <- rownames(meth_CD16Pos) #Apply the probe rownames to the 
filtered_dmt <- dmt.sites[rownames(dmt.sites) %in% pref_probes$IlmnID, ] #filter on only the preferred probes
filtered_dmt <- filtered_dmt[complete.cases(filtered_dmt), ] #remove any NA values
hyper_probes <- rownames(filtered_dmt[filtered_dmt$mean.diff > 0.2, ]) #Comparison is 15C vs 15D so 15C Beta minus 15D Beta
hypo_probes <- rownames(filtered_dmt[filtered_dmt$mean.diff < -0.2, ])
probe_anno_2 <- annotation(NK_RnB_subset2, type = "sites", add.names = T) #Call the annotation information for each probe from RnBeads
CD16Pos_hyper <- probe_anno_2[rownames(probe_anno_2) %in% hyper_probes, ] #Filter on just the hyper methylated probes
CD16Pos_hypo <- probe_anno_2[rownames(probe_anno_2) %in% hypo_probes, ] #Filter on just the hypo methylated probes

#Now for the Day14 cells
s.groups <- rnb.sample.groups(NK_RnB_subset3,"CellType")[[1]] #Pull comparison group
dmt.sites <- computeDiffTab.extended.site(meth(NK_RnB_subset3),s.groups[[1]],s.groups[[2]], paired = F) #Compute differential methylation in this condition for each probe
meth_D14 <- meth(NK_RnB_subset3, type = "sites", row.names = TRUE) #Get the beta values for the CD16- NK 
rownames(dmt.sites) <- rownames(meth_D14) #Apply the probe rownames to the 
filtered_dmt <- dmt.sites[rownames(dmt.sites) %in% pref_probes$IlmnID, ] #filter on only the preferred probes
filtered_dmt <- filtered_dmt[complete.cases(filtered_dmt), ] #remove any NA values
hyper_probes <- rownames(filtered_dmt[filtered_dmt$mean.diff > 0.2, ]) #Comparison is 15C vs 15D so 15C Beta minus 15D Beta
hypo_probes <- rownames(filtered_dmt[filtered_dmt$mean.diff < -0.2, ])
probe_anno_3 <- annotation(NK_RnB_subset3, type = "sites", add.names = T) #Call the annotation information for each probe from RnBeads
D14_hyper <- probe_anno_3[rownames(probe_anno_3) %in% hyper_probes, ] #Filter on just the hyper methylated probes
D14_hypo <- probe_anno_3[rownames(probe_anno_3) %in% hypo_probes, ] #Filter on just the hypo methylated probes

```

Function to write each foreground and background for de novo motif discovery through homer
using the findMotifsGenome.pl function through homer
```{r}
## ------------------------------
## Inputs
## ------------------------------
# All EPIC probes you tested
all_df <- probe_anno_2  

# Foreground sets you want to process
foregrounds <- list(
  D14_hyper     = D14_hyper,
  D14_hypo      = D14_hypo,
  CD16Neg_hyper = CD16Neg_hyper,
  CD16Neg_hypo  = CD16Neg_hypo,
  CD16Pos_hyper = CD16Pos_hyper,
  CD16Pos_hypo  = CD16Pos_hypo
)

## ------------------------------
## Helper functions
## ------------------------------

# Fix column names (in case of spaces)
rename_if_exists <- function(df, from, to) {
  if (from %in% colnames(df)) colnames(df)[colnames(df) == from] <- to
  df
}

make_windows <- function(df, flank = 100) {
  center <- floor((df$Start + df$End) / 2)
  ws <- center - flank
  we <- center + flank
  data.frame(
    Chromosome = df$Chromosome,
    Start0     = pmax(0L, as.integer(ws) - 1L),
    End1       = as.integer(we),
    Name       = df$ID,
    Score      = 0,
    Strand     = df$Strand,
    GC         = as.numeric(df$GC),
    CGI        = ifelse(is.na(df$CGI_Relation), "Unknown", df$CGI_Relation),
    stringsAsFactors = FALSE
  )
}

build_background <- function(fg_win, all_win, N_per = 10) {
  # Exclude fg probes
  pool <- subset(all_win, !(Name %in% fg_win$Name))
  
  # Bin by GC deciles
  gc_breaks <- quantile(pool$GC, probs = seq(0, 1, 0.1), na.rm = TRUE)
  pool$GCbin <- cut(pool$GC, breaks = gc_breaks, include.lowest = TRUE)
  fg_win$GCbin <- cut(fg_win$GC, breaks = gc_breaks, include.lowest = TRUE)
  
  pool$bin <- paste(pool$CGI, pool$GCbin, sep="|")
  fg_win$bin <- paste(fg_win$CGI, fg_win$GCbin, sep="|")
  
  sample_bin <- function(bin_label, need_n, pool_df) {
    cand <- which(pool_df$bin == bin_label)
    if (length(cand) >= need_n) {
      sample(cand, need_n, replace = FALSE)
    } else if (length(cand) > 0) {
      sample(cand, need_n, replace = TRUE)
    } else {
      # broaden to CGI only
      cgi <- strsplit(bin_label, "\\|")[[1]][1]
      cand2 <- which(pool_df$CGI == cgi)
      if (length(cand2) >= need_n) {
        sample(cand2, need_n, replace = FALSE)
      } else if (length(cand2) > 0) {
        sample(cand2, need_n, replace = TRUE)
      } else {
        sample(seq_len(nrow(pool_df)), need_n, replace = FALSE)
      }
    }
  }
  
  tab_fg <- table(fg_win$bin)
  idx_list <- integer(0)
  set.seed(42)
  for (b in names(tab_fg)) {
    need <- as.integer(tab_fg[[b]] * N_per)
    idx_list <- c(idx_list, sample_bin(b, need, pool))
  }
  
  bg_win <- pool[idx_list, , drop = FALSE]
  bg_win <- bg_win[!duplicated(bg_win[,c("Chromosome","Start0","End1","Name")]), ]
  list(fg_win = fg_win, bg_win = bg_win)
}

## ------------------------------
## Pre-clean full annotation
## ------------------------------
all_df <- rename_if_exists(all_df, "CGI Relation", "CGI_Relation")
keep_chr <- paste0("chr", c(1:22, "X", "Y"))
all_df <- subset(all_df, Chromosome %in% keep_chr & !is.na(Start) & !is.na(End))
all_win <- make_windows(all_df)

## ------------------------------
## Loop through foregrounds
## ------------------------------
for (nm in names(foregrounds)) {
  fg_df <- rename_if_exists(foregrounds[[nm]], "CGI Relation", "CGI_Relation")
  fg_df <- subset(fg_df, Chromosome %in% keep_chr & !is.na(Start) & !is.na(End))
  fg_win <- make_windows(fg_df)
  
  res <- build_background(fg_win, all_win, N_per = 10)
  
  # Write out to files
  write.table(res$fg_win[,c("Chromosome","Start0","End1")],
              file = paste0("fg_", nm, ".txt"),
              sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
  
  write.table(res$bg_win[,c("Chromosome","Start0","End1")],
              file = paste0("bg_", nm, ".txt"),
              sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
  
  cat("Wrote", nm, "foreground:", nrow(res$fg_win),
      "background:", nrow(res$bg_win), "\n")
}

```


